---
title: "GLM Model of Leach's Storm-Petrel Sexing"
author: "Autumn Pauly"
date: "12/31/2024"
output: html_document
---

Introduction

# Downloading Libraries
```{r install-tidyverse}
install.packages("tidyverse")
library(tidyverse)
```


# Reading the Data
```{r}
petrel_sexing <- read.csv("C:/Users/pauly/Desktop/summer_2024/greatduck_2024/autumn_2024/petrel_2024/r/petrel_morphometrics/petrelsex_john.csv")
```


### Identifying Means of the Sexes
This is where we are going to identify the means of the morphometric measurements of the  `female` and `male` individuals.

```{r means, warning=FALSE, echo=FALSE, message=FALSE}
# Calculate male and female means
means <- petrel_sexing %>%
  group_by(sex) %>%
  summarise(across(c(weight, chord, tl, hl, tu, cl, bdg, bdt), mean, na.rm = TRUE))

# Separate male and female means
male_means <- means %>% filter(sex == "male") %>% dplyr::select(-sex)
female_means <- means %>% filter(sex == "female") %>% dplyr::select(-sex)

# #printing the male and female means
# print(male_means)
# 
# print(female_means)
print(means)
```

# Univariate ANOVA to Compare Measurements Between the Sexes
We will perform ANOVA for each biometric measurement to compare means between males and females.

```{r anova}
# Perform ANOVA for each measurement
anova_results <- list()

measurements <- c("chord", "weight", "tl", "hl", "tu", "cl", "bdg", "bdt")

for (measure in measurements) {
  model <- aov(as.formula(paste(measure, "~ sex")), data = petrel_sexing)
  anova_results[[measure]] <- summary(model)
}

# Print results
anova_results
```

# Sexual Size Dimorphism (SSD) Index
The SSD index quantifies the size difference between males and females for each measurement as a percentage. The purpose of this is to measure and compare the extent of sexual dimorphism across traits in a standardized way. If males are larger on average, the SSD index will be positive. If females are larger, the SSD index will be negative.

```{r ssd}
# List of morphometric measurements
measurements <- c("weight", "chord", "tl", "hl", "tu", "cl", "bdt", "bdg")

# Calculate male and female means
means <- petrel_sexing %>%
  group_by(sex) %>%
  summarise(across(all_of(measurements), mean, na.rm = TRUE))

# Separate male and female means
male_means <- petrel_sexing %>%
  filter(sex == "male") %>%
  summarise(across(all_of(measurements), mean, na.rm = TRUE))

female_means <- petrel_sexing %>%
  filter(sex == "female") %>%
  summarise(across(all_of(measurements), mean, na.rm = TRUE))

# Calculate SSD
SSD <- (male_means / female_means - 1) * 100

# Add measurement names to the SSD result
SSD <- tibble::tibble(Measurement = measurements, SSD = as.numeric(SSD[1, ]))

# Print the SSD results
print(SSD)
```

# Generalized Linear Model (GLM) with Logistic Regression
A GLM is a flexible model for predicting a response variable (in this case, sex, a binary outcome: male or female) as a function of predictor variables (morphometric measurements like weight, chord, etc.).

The purpose of this model is to create a formula that predicts the sex of an individual based on its morphometric traits. The output is a probability that an individual belongs to one group (e.g., male), which can then be converted to a prediction (male or female). Predictor variables (traits) are tested for their significance in explaining sex differences.

### Prepare the Data
```{r sex_to_binary}
# Convert 'sex' to a binary factor if it's not already
petrel_sexing$sex <- as.factor(petrel_sexing$sex)
levels(petrel_sexing$sex) <- c("0", "1")  # 0 = female, 1 = male
```

### Fit the GLM with Logistic Regression
```{r initial_glm}
# Fit the logistic regression model
glm_model <- glm(sex ~ weight + chord + tl + hl + tu + cl + bdt + bdg,
                 data = petrel_sexing,
                 family = binomial(link = "logit"))

# Summary of the model
summary(glm_model)
```

### Evaluate Collinearity with Variance Inflation Factors (VIF)
 The variance inflation factor (VIF) is a measure of the amount of multicolinearity in regression analysis, essentially showing how much the variance of a regression coefficient is inflated due to correlations between independent variables in the model; a high VIF score suggests that a variable is highly correlated with other variables, potentially causing issues with model interpretation and stability. 
 
VIF helps identify which variables in a regression model are highly correlated with others, allowing you to address the issue by removing or transforming variables.
 
A VIF score above 5 is considered problematic, indicating a high level of multicollinearity. Scores closer to 1 suggest minimal collinearity. 


```{r vif_assessment}
# Install and load the 'car' package if necessary
if (!require(car)) install.packages("car")
library(car)

# Calculate VIFs
vif(glm_model)
```

### Model Selection using AIC
To identify the best combination of predictors, use the step() function to perform stepwise model selection based on the Akaike Information Criterion (AIC).

```{r model_selection_stepwise}
# Perform stepwise dplyr::selection
glm_step <- step(glm_model, direction = "both")

# Summary of the final model
summary(glm_step)
```

### Assess Original Model Accuracy
You can evaluate the model’s ability to correctly predict sex using cross-validation or by calculating metrics like accuracy, sensitivity, and specificity.

```{r original_model_accuracy}
# Predict probabilities
predicted_prob <- predict(glm_model, type = "response")

# Convert probabilities to binary predictions (0 or 1)
predicted_sex <- ifelse(predicted_prob > 0.5, 1, 0)

# Create a confusion matrix to compare predicted vs actual
confusion_matrix <- table(Predicted = predicted_sex, Actual = petrel_sexing$sex)

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))
```

### Assess Stepwise Model Accuracy
You can evaluate the model’s ability to correctly predict sex using cross-validation or by calculating metrics like accuracy, sensitivity, and specificity.

```{r stepwise_accuracy}
# Predict probabilities
predicted_prob <- predict(glm_step, type = "response")

# Convert probabilities to binary predictions (0 or 1)
predicted_sex <- ifelse(predicted_prob > 0.5, 1, 0)

# Create a confusion matrix to compare predicted vs actual
confusion_matrix <- table(Predicted = predicted_sex, Actual = petrel_sexing$sex)

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))
```
